---
title: "Assignment 1 - CSCI E-63C"
author: "Ethan Mallove"
date: "9/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

```{r problem1}
# different sample sizes we are going to try:
sample.sizes=c(3,10,50, 100, 500, 1000)

# we will use the vector below to save the standard deviations of the 
# *distribution of the means* at each given sample size.
# Note that it's ok to initialize with an empty vector of length 0 - if we index 
# it out-of-bounds while assigning a value later, the vector will autoexpand 
# on assignment, see examples in the slides) 
mean.sds = double(0) 

values = runif(1000)

i = 1
for ( N in sample.sizes ) { # try different sample sizes
  
  # INSERT YOUR CODE HERE: (you may want to check the slides). 
  
  # 1) At each given N (i.e. in each iteration of the outer loop) you have to draw large number 
  # (e.g. 1000) of samples of that size N, from the distribution of your choice (e.g. normal, 
  # uniform, exponential, ...), and calculate the mean of *each* of those samples. 
  # Save all those means into a vector 'm'. 
  #
  
  m <- NULL
  for (l in 1:N) {
    m[l] = mean(sample(values, N))
  }

  # 2) Now, with vector m in hand, we want to characterize how much the sample mean fluctuates
  # from one experiment (experiment=drawing a sample of N measurements) to the next. Instead of just
  # plotting a histogram as we did in class, this time we will calculate the standard 
  # deviation of the distribution represented by the vector m. Use function sd() to achieve that.
  #
  
  sd_ = sd(m)
  
  # 3) save the result (sd of the distributions of the means for current N) into the vector mean.sds 
  # defined above. You can use c() to concatenate, or you can use an indexing variable; 
  #  in the latter case you will need to add it to the code and increment properly
  
  mean.sds[i] = sd_
  
  i = i + 1
}

# at this point, you should have the vector mean.sds filled. It should have length 6 and keep the 
# calculated values of the standard deviation of the mean (known as the standard error of the mean, SEM) 
# at different sample sizes: mean.sds[1] is the SEM at N=3, mean.sds[2] is the SEM at N=10, and so on.

# let us now plot the SEM (i.e. the "typical" error we expect the sample mean to exhibit in any 
# given experiment) as a function of the sample size, N. 


```

## Plots

```{r SEM and sample sizes, echo=FALSE}
plot(sample.sizes, xlab="Sample Size", ylab="SEM", mean.sds, main="SEM vs sample size", pch=19)
lines(sample.sizes, 1/sqrt(sample.sizes), col='blue')
```


## Problem 2

```{r problem2}
N <- 1  # the number of i.i.d. variables X we are going to sum

# how many times we are going to repeat the "experiment" (see the text above for what we now call an experiment):
#n.repeats <- 1000
s.values <- numeric() # we will use this vector to store the value of the sum, s, in each experiment

for (n.repeats in c(1, 100, 500, 1000, 2000)) { # repeat the experiment 'n.repeats' times

   for (i.exp in 1:n.repeats) { # repeat the experiment 'n.repeats' times
   
      # More details below. In each "experiment" we must draw the values x1, ..., xN of all the 
      # random variables we are going to sum up:
     
      ### replace with correct call: 
      x <- runif(n.repeats)
   
     # the "measured" value of the random variable S in the current experiment is the sum of x1...xN;
     # calculate it and save into the vector s.values:
      
     ### replace with correct call: 
     s.values[i.exp] <- sum(x) # ...???...
   }

   hist(s.values, breaks = 20, main = paste("Histogram of sample size: " , n.repeats))

}
```

## Ending Questions

```{r ending question}
N <- 1  # the number of i.i.d. variables X we are going to sum

# how many times we are going to repeat the "experiment" (see the text above for what we now call an experiment):
#n.repeats <- 1000
s.values <- numeric() # we will use this vector to store the value of the sum, s, in each experiment

sds = numeric()

for (n.repeats in seq(1, 1000, 10)) { # repeat the experiment 'n.repeats' times

   for (i.exp in 1:n.repeats) {  # repeat the experiment 'n.repeats' times
   
      # More details below. In each "experiment" we must draw the values x1, ..., xN of all the 
      # random variables we are going to sum up:
     
      ### replace with correct call: 
      x <- runif(n.repeats)
   
     # the "measured" value of the random variable S in the current experiment is the sum of x1...xN;
     # calculate it and save into the vector s.values:
      
     ### replace with correct call: 
     s.values[i.exp] <- mean(x) # ...???...
     sd_ = sd(x)
   }

   sds[n.repeats] = sd_

   # hist(s.values, breaks = 20, main = paste("Histogram of sample size: " , n.repeats, "Std dev: ", sd_))

}

plot(sds)
# abline(lm(sds ~ seq(1, 1000, 10)))


```

> Answer the following question (5 points): suppose you have an arbitrary
> distribution and take a sample of N measurements from it. You calculate the
> mean of your sample. As we discussed, the sample mean is a random variable,
> of course. 

> How is the sample mean, as a random variable, distributed?

The sample mean is distributed normally.

> What is its (expected) mean (zero? infinity?  constant? which one if so)?

Constant

> What about standard deviation of the sample mean?  How does it behave as
> sample size N increases?


> Can anything be said about the shape of the distribution of sample means in
> the limit of large N? HINT: look at the definition of the sample mean!
